---
title: "Tuwel quiz score optimization"
author: "Emile Johnston"
output: pdf_document
toc: true
---

TUWEL enables the creation of quizzes with a special score system: for a given
batch of statements which must be marked as true or false, the score is 
calculated as follows: each true statement which is marked as true by the student
gives 1 point divided by the number of true statements in the batch, and each
false statement which is marked as true takes off 1 point, divided by the number
of false statements in the batch. Statements which are answered false don't 
result in positive or negative points.

# 1 with a "none of the above" box

In this first part, we are interested in the particular case where there are four statements per
batch, and there are five boxes to check: one for each statement, and one "none
of the above" box to check in case none of the statements are true. This is not
strictly necessary but is made so that it's impossible to get any points without
checking any boxes, and is a convenient way to give points when there are no 
true statements. It also introduces a bias in the scoring system: it artificially
increases the number of false "statements", therefore decreasing the potential 
penalty of guessing true compared to the potentially missed opportunity of 
guessing false. We will see that this results in a statistical benefit in 
guessing true in the majority of cases.

## 1.1. Putting the quiz logic into functions

### 1.1.1. Four statements, five boxes to check

First, we need to clearly distinguish the four statements, which are answered, 
and the five boxes, which are checked. Guessing is done on the statements, but
the score is determined from the checks.
This function takes the answers to the four questions, and returns the boxes that
would be checked according to these answers:

```{r}
n_q <- 4 # number of questions per batch

answers_to_checks <- function(answers) {
  # if all booleans in x are FALSE:
  if (!any(answers)) { 
    # if none of the statements are true, the last box "none of the above" is 
    # checked
    return(c(answers, TRUE))
  } # otherwise
  return(c(answers, FALSE)) 
  # if at least one of the statements is true, the last box is not checked
}

answers_to_checks(c(TRUE, FALSE, TRUE, FALSE))
answers_to_checks(c(FALSE, FALSE, FALSE, FALSE))
```


### 1.1.2. Calculating the score of a single statement
Next, a function that calculates the score contribution of a single question:

```{r}
question_score <- function(answer, correct_answer, n_true, n_false) {
  if (answer & correct_answer) { # statement is true, answer is TRUE
    return(1/n_true)
  }
  if (!answer) { # statement is true or false, answer is FALSE
    return(0)
  }
  return(-1/n_false) # statement is false, answer is TRUE
}

# generate all combinations for a boolean of length 2:
possibilities <- expand.grid(rep(list(c(TRUE, FALSE)), 2))

cat("For a case where there are 2 true statements and 2 false statements: \n")
for (i in 1:nrow(possibilities)) {
  cat("student's answer: ", possibilities[i,1],
      " correct answer: ", possibilities[i,2],
      "score: ", question_score(possibilities[i,1], possibilities[i,2], 2, 3), "\n")
}
```


### 1.1.3. Calculating the score of a batch of statements

Next, a function that calculates the score based on the boxes checked, and the correct
answers:

```{r}
score <- function(checks, correct_checks) {
  # count number of TRUEs in correct_checks
  num_true_statements <- sum(correct_checks)
  num_false_statements <- length(correct_checks) - num_true_statements
  scores <- numeric(length(checks))
  for (i in 1:length(checks)) {
    scores[i] <- question_score(checks[i], correct_checks[i], 
                                num_true_statements, num_false_statements)
  }
  sum(scores)
}

# randomly generate a boolean vector of length 4:
set.seed(1)
cat("answers to statements: ", 
    answers <- sample(c(TRUE, FALSE), n_q, replace = TRUE), "\n")
cat("boxes checked: ", checks <- answers_to_checks(answers), "\n")
cat("boxes that should be checked", 
    correct_checks <- c(TRUE, FALSE, TRUE, FALSE, FALSE), "\n")
cat("score: ", score(checks, correct_checks), "\n")
```


## 1.2. Expectancy of the score for completely random answers to all statements

What score do you get by answering everything randomly?

We can consider only the number of true or false statements in a batch without looking
at the order of the statements, because the truth value of the statements doesn't depend
on their order.

### 1.2.1. Generating random answers

First, generate all possible combinations of four booleans, where order does not matter:
```{r}
# create an empty boolean vector of lenght n_q:
booleans <- rep(FALSE, n_q) # n_q is the number of questions per batch
combinations <- rep(list(booleans), n_q+1)
for (i in 1:n_q) {
  combinations[[i+1]][1:i] <- TRUE
}
print(combinations)
cat("number of combinations: ", length(combinations))
```

```{r}
typeof(combinations[[1]])
```


### 1.2.2. Generating random statements

Next, generate all possible random answers: combinations where order does matter:
```{r}
permutations <- expand.grid(rep(list(c(TRUE, FALSE)), n_q))
permutations
```

```{r}
cat("typeof(permutations[[1]]) : " , typeof(permutations[[1]]), "\n")
cat("length(permutations) : ", length(permutations), "\n")
cat("nrows(permutations) : ", nrow(permutations), "\n")
cat(as.logical(permutations[1,]))
```

### 1.2.3. Calculating the score for all possible combinations of statements

for n questions, there are 2^n possible answers. Let's calculate the score for all of them:
```{r}
# This code chunk prints 330 lines. To make the pdf more readable the number of
# printed lines can be limited with a variable.
max_lines <- 40 # change this value to print more lines
end <- FALSE # to signal that the printed output ends
col_names <- c("0T 4F", "1T 3F", "2T 2F", "3T 1F", "4T 0F")
scores <- matrix(NA,
                 nrow=nrow(permutations),
                 ncol=length(combinations),
                 dimnames=list(NULL, col_names))
for (i in 1:length(combinations)) { # for each combination of true statements
  if (!end) {
    cat("\n if the correct answers are: ", combinations[[i]], "\n\n")
  }
  for (j in 1:nrow(permutations)) { # for each random answer
    s_answers <- as.logical(permutations[j,])
    s_checks <- answers_to_checks(s_answers)
    t_checks <- answers_to_checks(combinations[[i]])
    s <- score(s_checks, t_checks)
    scores[j,i] <- s
    if (max_lines > 0) {
      cat("student's answers: ", s_answers, "\n")
      cat("student checks these boxes: ", s_checks, "\n")
      # cat("correct answers: ", combinations[[i]], "\n")
      cat("boxes that should be checked: ", t_checks, "\n") # true checks
      cat("score: ", s, "\n")
      max_lines <- max_lines - 4
    }
    else {
      if (!end) {
        cat("\n Output is cut off here for readability. To print the full output, 
             change the max_lines variable to 350.")
        end <- TRUE
      }
    }
  }
}
```

Overall, the scores look like this:
```{r}
scores
```

Looking at the average per category:

```{r}
average_scores <- colMeans(scores)
average_scores
```

### 1.2.4. The score for a batch can't be negative

Now, we need to take into account the fact that a score for a batch of four statements
can't be negative. So we take the maximum of 0 and x for each score x:

```{r}
pos_scores <- pmax(scores, 0) # positive scores
pos_scores
```


### 1.2.5. Scores by random answers

Now let's calculate the average score to see the expectancy.
Per column:
```{r}
expectancy_per_trues <- colMeans(pos_scores)
expectancy_per_trues
```


### 1.2.6. Taking into account probabilities

Earlier we ignored the order of the statements, but we do need to take into account
the probability of each combination of true and false statements.
```{r}
probs <- choose(n_q,0:n_q) # probabilities of each combination of true and false statements
cat("probs: ", probs, "\n")
# normalise the probabilities:
probs <- probs/sum(probs)
cat("normalised probs: ", probs, "\n")
```

### 1.2.7. Overall expectancy

Now we can calculate the overall expectancy of answering randomly:
```{r}
overall_expectancy <- sum(expectancy_per_trues * probs)
overall_expectancy
```

## 1.3. Calculating all possible scenarios

### 1.3.1. Function for guessing the answers

This function takes answers to some of the four statements as input, and outputs 
the expectancy of the score for each possible guess for the remaining unanswered 
statements.  
For each case and for each possible guess, all possibilities for the remaining
uncertain statements are printed, along with the resulting score based on the
guess.  
For example if you know that one statement is false, and you guess TRUE FALSE
FALSE, and the outcome (actual truth value of the remaining statements) is TRUE
TRUE TRUE, the score for the batch is 0.33  
```{r}
guess <- function(params, nota_box=TRUE) {
  n_true <- params[1]
  n_false <- params[2]
  # n_true: number of statements in the batch which are already known to be true
  # n_false: number of statements in the batch which are already known to be false
  # nota_box: none of the above box, if it's added to the batch or not
  # Deal with irrelevant input:
  if (n_true < 0 | n_true > n_q | n_false < 0 | n_false > n_q) {
    stop("The number of true and false statements must be between 0 and", n_q)
  }
  if (n_true + n_false > n_q) {
    stop("There are only four statements per batch, so the number of true statements +
    false statements can't be more than", n_q)
  }
  if (n_true + n_false == n_q) {
    cat('You have already answered all four statements in this batch.')
    return(NULL)
  }

  # Explain the context:
  cat('You already know that', n_true, 'statements are true and', n_false,
  'statements are false. \n You have' , n_uncertain <- n_q - n_true - n_false,
      'statements left to answer, and there are', n_uncertain+1,
      "ways to guess them. \n\n")

  # Simulate all possible guesses:
  scores <- list()
  n_guesses <- n_q - n_true - n_false # number of remaining unanswered statements
  averages <- numeric() # the average score for each guess
  booleans <- rep(FALSE, n_guesses) # placeholders for the guesses
  guesses <- rep(list(booleans), n_guesses+1)
    for (i in 1:n_guesses) {
      guesses[[i+1]][1:i] <- TRUE
  }
  certain_answers <- c(rep(TRUE, n_true), rep(FALSE, n_false))
  permutations <- expand.grid(rep(list(c(TRUE, FALSE)), n_guesses)) # all possible cases
  for (guess in guesses) { # For each possible guess
    cat("if you guess", guess, ", these are the outcomes:", "\n")
    answers <- c(certain_answers, guess)
    if (nota_box) {
      checks <- answers_to_checks(answers)
    }
    if (!length(answers) == n_q) { # check if there are 4 answers
        cat('answers: ', answers, '\n')
        stop("The length of the answers vector is not equal to the number of questions per
         batch.")
    }
    guess_scores <- numeric(nrow(permutations))
    for (j in 1:nrow(permutations)) { # for each possible combination of correct answers
      correct_answers <- c(certain_answers, as.logical(permutations[j,]))
      if (!length(correct_answers) == n_q) { # check if there are 4 answers
        cat('correct answers: ', correct_answers, '\n')
        stop("The length of the correct answers vector is not equal to the number of questions per
         batch.")
      }
      if (nota_box) {
        correct_checks <- answers_to_checks(correct_answers)
        score <- pmax(score(checks, correct_checks), 0) # score for the batch must be positive
      } else {
        score <- pmax(score(answers, correct_answers), 0)
      }
      cat(as.logical(permutations[j,]),":", score, ", ")
      guess_scores[j] <- score
    }
    scores <- append(scores, list(guess_scores))
    cat("\n average: ", average <- mean(guess_scores), "\n\n")
    averages <- c(averages, average)
  }
  return (list(scores, guesses, averages))
}
```

### 1.3.2. Cases one by one

```{r}
cases <- matrix(c(0,0,  0,1,  0,2,  0,3,  1,0,  1,1,  1,2,  2,0,  2,1,  3,0),
                ncol=2, byrow=TRUE)
outcomes <- vector("list", nrow(cases))
```

#### 1.3.2.1. 0 true, 0 false statements
```{r}
outcomes[[1]] <- guess(cases[1,])
```


#### 1.3.2.2. 0 true, 1 false statement
```{r}
outcomes[[2]] <- guess(cases[2,])
```

#### 1.3.2.3. 0 true, 2 false statements
```{r}
outcomes[[3]] <- guess(cases[3,])
```

#### 1.3.2.4. 0 true, 3 false statements
```{r}
outcomes[[4]] <- guess(cases[4,])
```

#### 1.3.2.5. 1 true, 0 false statements
```{r}
outcomes[[5]] <- guess(cases[5,])
```

#### 1.3.2.6. 1 true, 1 false statement
```{r}
outcomes[[6]] <- guess(cases[6,])
```

#### 1.3.2.7. 1 true, 2 false statements
```{r}
outcomes[[7]] <- guess(cases[7,])
```

#### 1.3.2.8. 2 true, 0 false statements
```{r}
outcomes[[8]] <- guess(cases[8,])
```

#### 1.3.2.9. 2 true, 1 false statement
```{r}
outcomes[[9]] <- guess(cases[9,])
```

#### 1.3.2.10. 3 true, 0 false statements
```{r}
outcomes[[10]] <- guess(cases[10,])
```

### 1.3.3. Final results
Based on the outputs in the previous section:  
- if you know that one statement is true, and you don't know about the 3 others, you will
get the highest score expectancy by guessing 1 or 2 remaining statements to be true.  
- if you know that at least 2 statements are true, you will get the highest score
expectancy by guessing all remaining statements to be false.  
- in all other cases, you will get the highest score expectancy by guessing all remaining
statements to be true.  
  
Summary table:
```{r}
# make table with knitr kable:
pacman::p_load(knitr)
table <- data.frame(
  "What_you_already_answered" = c("0 T, 0 F", "0 T, 1 F", "0 T, 2 F",
                                   "0 T, 3 F", "1 T, 0 F", "1 T, 1 F",
                                   "1 T, 2 F", "2 T, 0 F", "2 T, 1 F",
                                   "3 T, 0 F"),
    "Best_guess" = c("TTTT", "TTT", "TT", "T or F", "TTF or TFF", "TT", "T", "FF",
                     "T or F", "F")
)
kable(table, format = "markdown", caption = "summary table")
```

# 2 without a "none of the above" box

Now we look at the case where there are still four statements to answer, but there is no
"none of the above" box.

## 2.1. Cases one by one

```{r}
outcomes_ <- vector("list", nrow(cases))
```

### 2.1.1. 0 true, 0 false statements
```{r}
outcomes_[[1]] <- guess(cases[1,], nota_box=FALSE)
```


### 2.1.2. 0 true, 1 false statement
```{r}
outcomes_[[2]] <- guess(cases[2,], nota_box=FALSE)
```

### 2.1.3. 0 true, 2 false statements
```{r}
outcomes_[[3]] <- guess(cases[3,], nota_box=FALSE)
```

### 2.1.4. 0 true, 3 false statements
```{r}
outcomes_[[4]] <- guess(cases[4,], nota_box=FALSE)
```

### 2.1.5. 1 true, 0 false statements
```{r}
outcomes_[[5]] <- guess(cases[5,], nota_box=FALSE)
```

### 2.1.6. 1 true, 1 false statement
```{r}
outcomes_[[6]] <- guess(cases[6,], nota_box=FALSE)
```

### 2.1.7. 1 true, 2 false statements
```{r}
outcomes_[[7]] <- guess(cases[7,], nota_box=FALSE)
```

### 2.1.8. 2 true, 0 false statements
```{r}
outcomes_[[8]] <- guess(cases[8,], nota_box=FALSE)
```

### 2.1.9. 2 true, 1 false statement
```{r}
outcomes_[[9]] <- guess(cases[9,], nota_box=FALSE)
```

### 2.1.10. 3 true, 0 false statements
```{r}
outcomes_[[10]] <- guess(cases[10,], nota_box=FALSE)
```

## 2.2 Conclusions

Let's compare the case where each batch of statements has a "none of the above"
box ("with_nota") and the case where it doesn't ("without_nota").
To keep only the most important information, we can make a function that 
extracts the best guess and score expectancy for each case:
```{r}
guess_to_string <- function(logical_vector) {
  char_vector <- ifelse(logical_vector, "T", "F")
  paste0(char_vector, collapse = "")
}

summarise <- function(cases, outcomes) {
  result <- data.frame("Known_answers" = rep(NA, length(outcomes)),
                       "Best_guess" = rep(NA, length(outcomes)),
                       "Score_expectancy" = rep(NA, length(outcomes)))
  result[,1] <- apply(cases, 1,
                      function(row) paste0(row[1], " T, ", row[2], " F"))
  for (i in 1:length(outcomes)) {
    idx <- which.max(outcomes[[i]][[3]]) # highest of averages
    result[i,2] <- guess_to_string(outcomes[[i]][[2]][[idx]])
    result[i,3] <- outcomes[[i]][[3]][idx]
  }
  return(result)
}
```

Collecting information for with_nota and without_nota respectively:
```{r}
summary <- summarise(cases, outcomes)
summary_ <- summarise(cases, outcomes_)
```

Comparing the two side by side:
```{r}
pacman::p_load(dplyr)
comparison <- summary %>% inner_join(summary_, by="Known_answers")
colnames(comparison) <- c("Known_answers", 
                          "Best_guess_with_nota", 
                          "Score_expectancy_with_nota",
                          "Best_guess_without_nota",
                          "Score_expectancy_without_nota")

comparison %>% select(Known_answers, 
                      Best_guess_with_nota, 
                      Best_guess_without_nota,
                      Score_expectancy_with_nota,
                      Score_expectancy_without_nota)
```

We can see that in 4 cases, the score expectancy is the same, and in the 
remaining 6 cases the score expectancy is higher when there is a NOTA box.  
This means that making a Tuwel quiz with NOTA boxes is inherently more favorable
to students (assuming that not everyone knows the answer to every statement).
Removing the NOTA box also changes the best guess in some cases: it is no longer
so biased towards guessing true.  
Overall, without the NOTA box, the best guess mostly consists in balancing out 
the true and false statements: when most known answers are true, it's best to 
guess false, and vice-versa.
